---
title: "Product-Level Profitability & Revenue Optimization Analysis"
subtitle: "Olist Brazilian E-Commerce Dataset"
author: "Mutkallam Warraich"
format:
  html:
    toc: true
    toc-depth: 2
    theme: cosmo
    number-sections: true
---

# Executive Summary

-   Total baseline profit was \$2.40M
-   A 10% reduction in freight costs increased profit by \$217K (+9.06%)
-   Profit is highly concentrated in a small number of categories (Pareto effect)
-   Several transactions operate at negative margins, indicating inefficiencies in pricing or cost structure

# Key Business Insights

-   A small group of product categories drives the majority of profit
-   Freight costs have a material impact on profitability
-   Low-margin and negative-margin orders present an opportunity for optimization
-   Pricing and cost control are the primary levers for improving profitability

# Recommendations

-   Optimize logistics to reduce freight costs (high ROI lever)
-   Identify and adjust pricing for low-margin or loss-making products
-   Focus inventory and marketing efforts on top-performing categories
-   Implement margin monitoring at the SKU level

# Code

## Data Preparation

This section loads and prepares the datasets used for analysis.

```{r}
library(tidyverse)
library(janitor)
library(lubridate)

# Load datasets
order_items <- read_csv("data/raw/olist_order_items_dataset.csv") |> clean_names()
orders      <- read_csv("data/raw/olist_orders_dataset.csv") |> clean_names()
products    <- read_csv("data/raw/olist_products_dataset.csv") |> clean_names()
customers   <- read_csv("data/raw/olist_customers_dataset.csv") |> clean_names()
```

```{r}
names(order_items)
names(orders)
names(products)
names(customers)
```

## Data Integration

This step combines multiple datasets into a unified order-level dataset for analysis.

```{r}
master_orders <- order_items %>%
  left_join(
    orders %>%
      select(
        order_id,
        customer_id,
        order_status,
        order_purchase_timestamp,
        order_delivered_customer_date
      ),
    by = "order_id"
  ) %>%
  left_join(
    products %>%
      select(
        product_id,
        product_category_name
      ),
    by = "product_id"
  ) %>%
  left_join(
    customers %>%
      select(
        customer_id,
        customer_city,
        customer_state
      ),
    by = "customer_id"
  )
```

```{r}
glimpse(master_orders)
```

## Data Cleaning

This step filters the dataset to include only valid completed transactions.

```{r}
master_orders_clean <- master_orders %>%
  filter(order_status == "delivered") %>%
  filter(!is.na(price), !is.na(freight_value), !is.na(product_category_name))

glimpse(master_orders_clean)
```

## Profitability Framework

These will have all the key financial metrics including revenue, cost, profit, and margin being calculated.

```{r}
master_orders_clean <- master_orders_clean %>%
  mutate(
    revenue = price,
    total_cost = 0.65 * price + freight_value,   # assumed 65% cost
    gross_profit = revenue - total_cost,
    margin_pct = gross_profit / revenue
  )

glimpse(master_orders_clean)
```

## Category-Level Performance

This section summarizes revenue and profitability across product categories.

```{r}
category_kpis <- master_orders_clean %>%
  group_by(product_category_name) %>%
  summarise(
    total_revenue = sum(revenue, na.rm = TRUE),
    total_profit = sum(gross_profit, na.rm = TRUE),
    avg_margin = mean(margin_pct, na.rm = TRUE),
    total_orders = n()
  ) %>%
  arrange(desc(total_profit))

category_kpis
```

### Top and Bottom Categories

This section identifies the highest and lowest performing product categories.

```{r}
top_categories <- category_kpis %>%
  slice_max(total_profit, n = 10)

top_categories
```

```{r}
bottom_categories <- category_kpis %>%
  slice_min(total_profit, n = 10)

bottom_categories
```

## Pareto Analysis

This analysis examines how profit is concentrated across categories.

```{r}
pareto <- category_kpis %>%
  mutate(
    profit_share = total_profit / sum(total_profit),
    cumulative_profit = cumsum(profit_share)
  )

pareto
```

### Pareto Visualization

```{r}
pareto_top <- pareto %>%
  slice_max(total_profit, n = 15)

ggplot(pareto_top, aes(x = reorder(product_category_name, total_profit), y = cumulative_profit)) +
  geom_line(group = 1) +
  geom_point() +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Pareto (Top 15 Categories)",
    x = "Category",
    y = "Cumulative Profit %"
  )
```

## Price vs Profit Analysis

This section explores the relationship between pricing and profitability.

```{r}
ggplot(master_orders_clean, aes(x = price, y = gross_profit)) +
  geom_point(alpha = 0.3) +
  labs(
    title = "Profit vs Price (with Freight)",
    x = "Price",
    y = "Gross Profit"
  )
```

## Margin Distribution

This section shows how profit margins are distributed across transactions.

```{r}
ggplot(master_orders_clean, aes(x = margin_pct)) +
  geom_histogram(bins = 50) +
  labs(
    title = "Margin Distribution",
    x = "Margin %",
    y = "Count"
  )
```

## Loss Analysis

This section identifies transactions generating negative profit.

```{r}
loss_products <- master_orders_clean %>%
  filter(gross_profit < 0)

nrow(loss_products)
```

## Scenario Analysis

This section simulates the impact of reducing freight costs on profitability.

```{r}
scenario <- master_orders_clean %>%
  mutate(
    new_cost = 0.65 * price + (freight_value * 0.9),
    new_profit = price - new_cost
  )

# Compare total profit
baseline_profit <- sum(master_orders_clean$gross_profit)
new_profit <- sum(scenario$new_profit)

profit_lift <- new_profit - baseline_profit
profit_lift
```

```{r}
profit_lift_pct <- profit_lift / baseline_profit

profit_lift_pct
```

## Financial Impact Summary

```{r}
tibble(
  baseline_profit = baseline_profit,
  new_profit = new_profit,
  profit_lift = profit_lift,
  profit_lift_pct = profit_lift_pct
)
```
